---
title: "Fundamentals of Computing and Data Display"
author: "Robert A. Schultz"
subtitle: Surv. 727
output:
  word_document: default
  html_notebook: default
  pdf_document: default
---

## Setup

```{r}
suppressPackageStartupMessages({
library(tidyverse)
library(DBI)
library(dbplyr)
library(bigrquery)
})
```

## Database connection

In this notebook we --again -- use Google BigQuery, "Google's fully managed, petabyte scale, low cost analytics data warehouse". Instruction on how to connect to Google BigQuery can be found here:

https://db.rstudio.com/databases/big-query/

After following the steps described on this website and initializing a project, paste your project ID into the following chunk.

```{r}
project <- "hw-295220"
```

First, set up the connection to the database. This time we use the Chicago crime database, which is a BigQuery version of the Chicago crime API we used in earlier classes.

```{r}
con <- dbConnect(
  bigrquery::bigquery(),
  project = "bigquery-public-data",
  dataset = "chicago_crime",
  billing = project
)
con 
```

Show tables that are available.

```{r}
dbListTables(con)
```

## SQL and dbplyr

Information on the `crime` table can be found here:

https://cloud.google.com/bigquery/public-data/chicago-crime-data

Write a first query that counts the number of rows of the `crime` table in the year 2016. The following code chunks expect SQL code.

#1000 rows

```{sql connection= con}
SELECT *  
FROM crime 
WHERE year=2016
```


Next, count the number of arrests grouped by `primary_type` in 2016. Note that is a somewhat similar task as above, with some adjustments on which rows should be considered. Sort the results, i.e. list the number of arrests in a descending order.

```{sql connection= con}
SELECT primary_type, count(arrest)
FROM crime
WHERE year = 2016 AND arrest = TRUE
GROUP BY primary_type
ORDER BY COUNT(*) DESC
```

We can also use the `date` for grouping. Count the number of arrests grouped by hour of the day in 2016. You can extract the latter information from `date` via `EXTRACT(HOUR FROM date)`. Which time of the day is associated with the most arrests? 

#Appears to be in military or 24 hour time, I know here stateside we just do AM/PM. Appears to be 7pm??

```{sql connection= con}
SELECT EXTRACT(HOUR FROM date), count(arrest)
FROM crime
WHERE year = 2016 and arrest = TRUE
GROUP BY EXTRACT(HOUR FROM date)
ORDER BY COUNT(*) DESC
```

Focus only on `HOMICIDE` and count the number of arrests for this incident type, grouped by year. List the results in descending order.

```{sql connection= con}
SELECT year, COUNT(*) AS homicide_arrest
FROM crime
WHERE primary_type = 'HOMICIDE' and arrest = TRUE
GROUP BY year
ORDER BY COUNT(*) DESC

```

Find out which districts have the highest numbers of arrests in 2015 and 2016. That is, count the number of arrests in 2015 and 2016, grouped by year and district. Again, list the results in descending order (within year).

```{sql connection= con}
SELECT district, year, COUNT (arrest)
FROM crime
WHERE year IN (2015, 2016)
GROUP BY district, year
ORDER BY COUNT(*) DESC

```

Now, try to compute the difference between the number of arrests in 2016 and 2015 by district. Order the results such that the district with the highest decrease in arrests comes first.

```{sql connection= con}
SELECT district,
 COUNTIF(year = 2015) AS arrests_2015,
 COUNTIF(year = 2016) AS arrests_2016,
 COUNTIF(year = 2016) - COUNTIF(year = 2015) AS change
FROM crime
WHERE arrest = TRUE
Group By district
ORDER BY change ASC
```

Lets switch to writing queries from within R via the `DBI` package. Create a query object that counts the number of arrests grouped by `primary_type` of district 11 in year 2016. The results should be displayed in descending order.

```{r}
sql2 <- "SELECT primary_type, COUNT(*) AS arrest_count
FROM crime
WHERE year = 2016 AND district = 11 AND arrest = TRUE
GROUP BY primary_type
ORDER BY COUNT(*) DESC"

```

Execute the query.

```{r}
dbGetQuery(con, sql2)
```

Try to write the very same query, now using the `dbplyr` package. For this, you need to first map the `crime` table to a tibble object in R.

```{r}
CRIME <- tbl(con, "crime")
str(CRIME)
class(CRIME)
```

Again, count the number of arrests grouped by `primary_type` of district 11 in year 2016, now using `dplyr` syntax.

```{r}
CRIME %>%
  filter(year == 2016) %>%
  filter(district == 11) %>%
  filter(arrest == TRUE) %>%
  group_by(primary_type) %>%
  summarise(total=n()) %>%
  arrange(desc(total))

```

Count the number of arrests grouped by `primary_type` and `year`, still only for district 11. Arrange the result by `year`.

```{r}
CRIME %>%
  filter(district == 11) %>%
  filter(arrest == TRUE) %>%
  group_by(primary_type, year) %>%
  summarise(total=n()) %>%
  arrange(year)
```

Assign the results of the query above to a local R object.

```{r}
Crime_data <-
  CRIME %>%
  filter(district == 11) %>%
  filter(arrest == TRUE) %>%
  group_by(primary_type, year) %>%
  summarise(total=n()) %>%
  arrange(year) %>%
  collect()
```

Confirm that you pulled the data to the local environment by displaying the first ten rows of the saved data set.

```{r}
str(Crime_data)
head(Crime_data, 10)
```

Close the connection.

```{r}
dbDisconnect(con)
```
